---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data
1. Load the data.
```{r chunk1}
unzip("activity.zip")
mydt <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
1. Calculate the total number of steps per day.
```{r chunk2.1}
sumdt <- mydt %>% group_by(date) %>% summarize(Total = sum(steps, na.rm = T))
```

2. Make a histogram
```{r chunk2.2}
qplot(sumdt$Total, xlab = "Total Daily Steps", bins = 10)
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r chunk2.3}
mymean <- round(mean(sumdt$Total),3)
mymedian <- median(sumdt$Total)
```

The mean of the total daily steps equals `r mymean` and the median equals `r mymedian`.


## What is the average daily activity pattern?
1. Time series plot of average steps per each 5 minute interval.
```{r chunk3.1}
avedt <- mydt %>% group_by(interval) %>% summarize(mean = mean(steps, na.rm = T), sd = sd(steps, na.rm = T))
qplot(x = avedt$interval, y = avedt$mean, geom = "line", xlab = "5 Minute Interval", ylab = "Mean Steps")
```

2. Find the interval with the maximum average of steps.
```{r chunk3.2}
mymax <- max(avedt$mean)
myintv <- avedt[avedt$mean == mymax,]$interval
```

The 5 minute interval which contains the maximum average steps is `r myintv`.


## Imputing missing values
1. Calculate and report the total number of missing values in the dataset.
```{r chunk4.1}
myNA <- sum(is.na(mydt$steps))
```

The total number of missing values equals `r myNA`.


2. Devise a strategy for filling in all of the missing values in the dataset.

I decided to use the mean and SD of the corrosponding 5 minute interval to create a normal distribution from which to sample a random variable. If we generate a negative number, we will use 0 instead. While this might be too complicated, I thought it would be fun to code.
```{r chunk4.2, cache=TRUE}
mydt2 <- mydt
set.seed(123)
for (i in 1:nrow(mydt2)){
    if (is.na(mydt2[i,]$steps)){
        mydt2[[i,1]] <- rnorm(1, mean = avedt[avedt$interval == mydt2[i,]$interval,]$mean,
                        sd = avedt[avedt$interval == mydt2[i,]$interval,]$sd)
    }
}
mydt2[mydt2$steps < 0,]$steps <- 0

```

3. Make a histogram of the new data and compare to the old data.
```{r chunk4.3}
sumdt2 <- mydt2 %>% group_by(date) %>% summarize(Total = sum(steps, na.rm = T))
qplot(sumdt2$Total, xlab = "Total Daily Steps", bins = 10)

mymean2 <- round(mean(sumdt2$Total),3)
mymedian2 <- median(sumdt2$Total)

mymeanD <- abs(mymean - mymean2)
mymedianD <- abs(mymedian - mymedian2)
```

The mean of the total daily steps in the new data equals `r as.character(mymean2)`, which is different from the old data by `r mymeanD`. The median of the total daily steps in the new data equals `r as.character(mymedian2)`, which is different from the old data by `r mymedianD`.


## Are there differences in activity patterns between weekdays and weekends?
1. Create new factor, 'weekday' or 'weekend'.
```{r chunk5.1}
mydt2$date <- as.Date(mydt2$date)
mydt2$daytype <- weekdays(mydt2$date)
wknd <- c("Saturday", "Sunday")
mydt2[mydt2$daytype %in% wknd,]$daytype <- "Weekend"
mydt2[!mydt2$daytype == "Weekend",]$daytype <- "Weekday"
mydt2$daytype <- as.factor(mydt2$daytype)
```

2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
```{r chunk5.2}
avedt2 <- mydt2 %>% group_by(daytype, interval) %>% summarize(mean = mean(steps, na.rm = T), .groups = "keep")
qplot(data = avedt2, x = interval, y = mean, facets = daytype ~ ., geom = "line",
      xlab = "5 Minute Interval", ylab = "Average Steps")
```


